---
title: "Assessment of AMR-Associated Indicator Pathogens and Environmental AMR Risk in Irish Bathing Waters"
author: "Brian O'Sullivan"
date: "June, 2026"
output:
  html_document:
    toc: true
---

# Environmental AMR risk assessment pipeline
AMR risk assessment of inland and coastal waters using source attribution and source-specific AMR risk weighting of routine bathing water microbiological monitoring data.

## Load the necessary datasets.
The script `primaryAmrDriverDataDownload.R` loads the data objects listed below. Any required source files that are not already present in the `DATA` directory are downloaded automatically.

This analysis pipeline requires the file `DATA/epaBqXyrsMedians.tsv`, which contains six-year median pathogen levels and SOM classifications for all national bathing water quality monitoring sites. This file is generated by `SomBathingWaterQualityAnalysis.Rmd`. If you have not already run that markdown document, please do so before continuing.


| R Object Name | Dataset Description | Provider | Data Update Frequency | Analysis Period | Licence |
|:------------------------|:----------|:----------|:----------------------|:--------------------------|:----------|
| edSfAVA42 | Livestock | CSO | 1 in 10 years |2020 |  CC BY 4.0^a^ |
| edSfSAP2022T6T7ED_2022 | Domestic wastewater | CSO | 1 in 5 years | 2022 | CC BY 4.0^a^ |
| wwEp | Urban wastewater plants | EPA | Single snapshot^b^ | 2025 | CC BY 4.0^a^ |
| wwtpFacilities | Urban wastewater emissions | EPA | Single snapshot^b^ | 2025 | CC BY 4.0^a^ |
| ielFacilities | Industrial emissions licences | EPA | Single snapshot^b^ | 2025 | CC BY 4.0^a^ |
| ielAquaculture | Aquaculture | DAFM | Single snapshot^b^ | 2025 | Open data |
| hiqaNursingHomes & hospitalsSf | Healthcare | HIQA & OSM | Single snapshot^b^ | 2024 (HIQA) and 2025 (OSM) | Open data |
| bqDataComplete | Bathing water quality | EPA | 13 times per month (in season^c^) | 2020 to 2025 | CC BY 4.0^a^ |

^a^ CC BY 4.0 indicates data released under the Creative Commons Attribution 4.0 International licence.
^b^ Single snapshot indicates that only current data were available and no historical time series existed.
^c^ Bathing season defined as 1 June-15 September.


```{r}
# Specify the time frame for bathing water quality assessment.
# The period includes data from the start year up to and including the end year.
# Modify this range and re-run the analysis as required.
yearStart=2020
yearEnd=2025

source("primaryAmrDriverDataDownload.R")

# This table was created by running SomBathingWaterQualityAnalysis.Rmd
# Make sure required data exists.
if (!file.exists("DATA/epaBqXyrsMedians.tsv")) {
  # Then download it
  message("Please create the file \"DATA/epaBqXyrsMedians.tsv\" as described above.")
  knit_exit()
}

epaBqXyrsMediansStaged=read.table(file='DATA/epaBqXyrsMedians.tsv', sep='\t', header = TRUE)
```
## Coastal risk assessment

### Define Shoreline Segments
Divide Irish coastline into discrete shoreline segments based on monitoring coverage and contamination status.

```{r, fig.height = 8, fig.width = 7.5, fig.align = "left"}

# Subset to coastal bathing waters.

tCoastalStations = epaBqXyrsMediansStaged[-match(inlandBathingWaters, epaBqXyrsMediansStaged$Code, nomatch=0),]

tCoastalStations = st_as_sf(tCoastalStations,
                            coords = c("Easting", "Northing"),
                            crs = 29903)

# Add index of nearest coastline vertex on the map of Ireland to each coastal test station.
# We will use this to get the coastling distance between water monitoring stations in our analysis.

# First transform Ireland coastline map to ITM
irlMainland = st_transform(repIrl, crs=29903)

# Make it a line, instead of a multiline
irlMainland = st_cast(irlMainland,"LINESTRING")

# Its a complex coastline with over small islands.
# Simplify to ease processing etc.
# Exclude any islands for now, just use mainland Ireland as approximation.
irlCoastlineAprox=irlMainland[irlMainland$Shape__Length>10000,]

# Split out Achill island, we'll join this to the coastline later
achill=as.matrix(rbind(as.data.frame(st_coordinates(irlCoastlineAprox[17,])),
                       as.data.frame(st_coordinates(irlCoastlineAprox[15,])),
                       as.data.frame(st_coordinates(irlCoastlineAprox[25,]))))
achill =   st_sfc(st_linestring(achill))
st_crs(achill) <- 29903 

# Join the main sections of coastline from whats left.
# Set up coordinates for joined coastline.
irlCoastlineAprox = as.matrix(rbind(as.data.frame(st_coordinates(irlCoastlineAprox[33,])),
                                    as.data.frame(st_coordinates(irlCoastlineAprox[34,])),
                                    as.data.frame(st_coordinates(irlCoastlineAprox[1,]))))



irlCoastlineAprox =   st_sfc(st_linestring(irlCoastlineAprox))
st_crs(irlCoastlineAprox) <- 29903

irlCoastlineAproxMinusAchill = irlCoastlineAprox


# Now join Achill to the mainland. The seperation between Achill and the mainland is very small.
# There are also 3 test stations on Achill.
# It is a reasonable approximation to make.
achillPoints=st_cast(achill,"POINT")

# Find where it "joins" the mainland
pt=unlist(st_intersects(st_nearest_points(irlCoastlineAprox,achill),achillPoints))

# Reorder points on Achill, (clockwise from the join with mainland.)
test=achillPoints[c(pt:1,(pt+2):length(achillPoints))]

# Find where the mainland "joins" Achill.
mainlandPoints=st_cast(irlCoastlineAprox,"POINT")
pt=unlist(st_intersects(st_nearest_points(irlCoastlineAprox,achill),mainlandPoints))

newMain = rbind(as.data.frame(st_coordinates(mainlandPoints[1:pt])),
                as.data.frame(st_coordinates(test)))
newMain = rbind(newMain,as.data.frame(st_coordinates(mainlandPoints[pt:length(mainlandPoints)])))
newMain = as.matrix(newMain)
newMain =   st_sfc(st_linestring(newMain), crs=29903)  

irlCoastlineAprox = newMain

# Get all the vertices (points) that make up sections of this line.
irlVertices = st_cast(irlCoastlineAprox,"POINT")

# Drop z dimension
irlVertices = st_zm(irlVertices, drop = TRUE, what = "ZM")

# Compute the distance between each vertex.
distanceToLastVertex = c(0, st_distance(
  irlVertices[1:length(irlVertices)-1,],
  irlVertices[2:length(irlVertices),],
  by_element = TRUE
))

# Get the cumulative sum to provide a vector with distance of each vertex along the coastline from the start. 
distanceToStartOfCoastline = cumsum(distanceToLastVertex)

# Create an sf object containing the specified parameters baked in.
irlVertices =  st_as_sf(data.frame(distanceToStartOfCoastline = distanceToStartOfCoastline,
                                   distanceToLastVertex = distanceToLastVertex,
                                   geometry = st_geometry(irlVertices)))

# Create a rough polygon of republic.
irlPoly=st_cast(irlCoastlineAproxMinusAchill,"POLYGON")

# Find the nearest vertex on the coastline for each of the bathing water quality test stations
# and append the required distances to the 5 year median test station data

tCoastalStations$nearestPointsIndices <- st_nearest_feature(tCoastalStations, irlVertices)


# Also add distance to the start of the coastline for each of the coastal stations.
tCoastalStations$distanceToStartOfCoastline = irlVertices$distanceToStartOfCoastline[tCoastalStations$nearestPointsIndices]

# Sort all coastal stations in order as you move around the coastline of Ireland.
tCoastalStations = tCoastalStations[order(tCoastalStations$distanceToStartOfCoastline),]

#########COASTAL SECTIONS WITH NO BAHTING WATER QUALITY MONITORING SITES###################
tmpInterStationSections = data.frame(start= c(0,tCoastalStations[1:(length(tCoastalStations$nearestPointsIndices)-1),]$nearestPointsIndices),
                                     end=tCoastalStations[1:length(tCoastalStations$nearestPointsIndices),]$nearestPointsIndices
)

# 
tmpInterStationDistances=apply(tmpInterStationSections[, c("start", "end")], 1, function(x) {x[1]=x[1]+1;sum(irlVertices$distanceToLastVertex[x[1]:x[2]])})

# Find out the last stations before a section of no coverage.
# A shoreline sections with no monitoring site and exceeding 120 km in length, which is approximately
# 2% of the total coastline length, as defined by the
# Tailte Éireann Coast – National 250K Map of Ireland are designated as no coverage.
noCoverage=which(tmpInterStationDistances> 120000)

# Extend a coverage 20km past the last station before a no coverage section.
noCoverageThreshold=20000

# End of coverage lines
noCoverageMultiline = c()

# Add entries in the coastal monitoring stations object collection to indicate where no coverage sections start / end.
for(j in noCoverage) {
  
  strtIdx=tCoastalStations[j-1,]$nearestPointsIndices
  endIdx=tCoastalStations[j,]$nearestPointsIndices
  
  i=strtIdx
  noCoverageStartDist=irlVertices[i,]$distanceToStartOfCoastline+noCoverageThreshold
  
  # Move forward around the coastline until you exceed this distance
  # then plant a no coverage marker at that location.    
  while(irlVertices[i,]$distanceToStartOfCoastline<noCoverageStartDist)
    i=i+1;
  
  # Create new out of coverage point to add to tCoastalStations    
  noCovStartPt=data.frame(
    geometry=st_sfc(st_point(st_coordinates(irlVertices[i,]))),
    LocationId=NA,
    Code=NA,
    Name=NA,
    CountyCode=NA,
    CountyName=NA,
    OrganisationName=NA,
    EdenMonitoredStationCode=NA,
    EcoliResult=NA,
    EnterococciResult=NA,
    EcoliTwsmmResult=NA,
    EnterococciTwsmmResult=NA,
    class=NA,
    ecoliEnterococciCluster=NA,
    pathogenSeverityLevel=NA,
    nearestPointsIndices=i,
    distanceToStartOfCoastline=irlVertices[i,]$distanceToStartOfCoastline)
  
  noCovStartPt=st_as_sf(noCovStartPt, crs = 29903)
  
  # Add this out of coverage point  
  tCoastalStations=rbind(tCoastalStations,noCovStartPt)
  
  # Now see where we come back into coverage again
  noCoverageStartDist=irlVertices[endIdx,]$distanceToStartOfCoastline-noCoverageThreshold
  
  # Move forward around the coastline until you exceed this distance
  # then plant a no coverage marker at that location.
  while(irlVertices[i,]$distanceToStartOfCoastline<noCoverageStartDist)
    i=i+1;
  
  # Create new out of coverage point to add to tCoastalStations    
  noCovEndPt=data.frame(
    geometry=st_sfc(st_point(st_coordinates(irlVertices[i,]))),
    LocationId=NA,
    Code=NA,
    Name=NA,
    CountyCode=NA,
    CountyName=NA,
    OrganisationName=NA,
    EdenMonitoredStationCode=NA,
    EcoliResult=NA,
    EnterococciResult=NA,
    EcoliTwsmmResult=NA,
    EnterococciTwsmmResult=NA,
    class=NA,
    ecoliEnterococciCluster=NA,
    pathogenSeverityLevel=NA,
    nearestPointsIndices=i,
    distanceToStartOfCoastline=irlVertices[i,]$distanceToStartOfCoastline)
  
  noCovEndPt=st_as_sf(noCovEndPt, crs = 29903)
  
  # Add this out of coverage point  
  tCoastalStations=rbind(tCoastalStations,noCovEndPt)  
  
  coastlineOfInterest <- st_sfc(st_linestring(st_coordinates(irlVertices[noCovStartPt$nearestPointsIndices:noCovEndPt$nearestPointsIndices,])))
  
  st_crs(coastlineOfInterest) <- 29903
  
  noCoverageMultiline = rbind(noCoverageMultiline,coastlineOfInterest)
  
}

# Re-order the water quality monitoring stations data frame in coastline order.
tCoastalStations = tCoastalStations[order(tCoastalStations$distanceToStartOfCoastline),]

# Use this to create an sf object
# Also tag it with highest pathogen severity station in this group.
noCoverageMultiline = st_as_sf(data.frame(noCoverageMultiline),
                               crs = 29903)


# Now deal with coastline stretches marked by above ambient pathogen levels
# Draw up a list of monitoring sites that are above ambient level (probList below)
highPathogenStations=which(tCoastalStations$pathogenSeverityLevel !=1 & !is.na(tCoastalStations$pathogenSeverityLevel))
probList=unname(split(highPathogenStations, highPathogenStations - seq_along(highPathogenStations)))


riskCollection = c()
riskLinesCollection = c()
alreadyAssigned = c()

for(i in probList)
{
  lIndex=unlist(i)
  
  # Expand to include ambient or out of coverage station at either side of the coastal segment.
  
  # Cut out this section of coastline upto and including the next ambient or out of coverage site at
  # either end.
  coastStart=tCoastalStations[min(lIndex)-1,]$nearestPointsIndices
  coastStop=tCoastalStations[max(lIndex)+1,]$nearestPointsIndices
  
  riskZone = st_sfc(st_linestring(st_coordinates(irlVertices[coastStart:coastStop,])), crs = 29903)
  riskZone <- st_sf(geometry = riskZone)
  
  # Now record the line of risk along the coast, we will use this to get the surrounding
  # cattle densities etc., later
  # Also name this section of coastline after the bathing area with the highest pathogen levels
  riskZone$Name=tCoastalStations[lIndex,][rev(order(tCoastalStations[lIndex,]$pathogenSeverityLevel)),]$Name[1]
  riskZone$maxPathogenSeverityLevel=unlist(max(tCoastalStations[lIndex,]$pathogenSeverityLevel))
  lIndex = (min(lIndex)-1):(max(lIndex)+1)
  riskZone$LocationId=list(tCoastalStations[lIndex,]$LocationId)
  riskZone$distanceToStartOfCoastline=list(tCoastalStations[lIndex,]$distanceToStartOfCoastline)
  riskLinesCollection = rbind(riskLinesCollection,
                              riskZone)
  
  
  # Expand risk zone 2km out to sea.
  riskZone = st_buffer(riskZone, 2000)
  
  # Add it to the risk collection. Also tag it with highest pathogen severity station in this group.
  riskCollection = rbind(riskCollection,
                         cbind(riskZone,
                               maxPathogenSeverityLevel=unlist(max(tCoastalStations[lIndex,]$pathogenSeverityLevel))))
  
  # Remove stations included in this segment from the list...
  # Maintain a list of all stations already assigned to above ambient coastal segments so
  # we can exclude these from further segment processing.
  alreadyAssigned = c(alreadyAssigned,lIndex)
}
alreadyAssigned=unique(alreadyAssigned)

# Now deal with coastline stretches marked by background/ambient pathogen levels

# First deal with coastline stretches dealing with groups of adjacent ambient level stations.
# Store them in a list with each element containing a station or adjacent stations.
lowPathogenStations=which(tCoastalStations$pathogenSeverityLevel==1  & !is.na(tCoastalStations$pathogenSeverityLevel))
noProbList=unname(split(lowPathogenStations, lowPathogenStations - seq_along(lowPathogenStations)))

for(i in noProbList[lengths(noProbList)>1])
{
  lIndex=unlist(i)
  
  start=min(lIndex)
  stop=max(lIndex)
  
  # Get a name for this coastline section.
  tName=tCoastalStations[start+ceiling((stop-start)/2),]$Name
  
  # if the next station either side of the start/stop points is a no coverage marker then continue the line as far as that.
  if((start-1)>0 && is.na(tCoastalStations[(start-1),]$pathogenSeverityLevel))
  {
    start=start-1
  }
  
  if((stop+1)<nrow(tCoastalStations) && is.na(tCoastalStations[(start+1),]$pathogenSeverityLevel))
  {
    stop=stop+1
  }
  
  # Cut out this section of coastline
  coastStart=tCoastalStations[start,]$nearestPointsIndices
  coastStop=tCoastalStations[stop,]$nearestPointsIndices
  
  riskZone = st_sfc(st_linestring(st_coordinates(irlVertices[coastStart:coastStop,])), crs = 29903)
  riskZone <- st_sf(geometry = riskZone)
  
  # Now record the line of risk along the coast, we will use this to get the surrounding
  # cattle densities etc., later
  # Also tag it with a name of the station in the middle of this group.
  riskZone$Name=tName
  riskZone$maxPathogenSeverityLevel=1
  riskZone$LocationId=list(tCoastalStations[start:stop,]$LocationId)
  riskZone$distanceToStartOfCoastline=list(tCoastalStations[start:stop,]$distanceToStartOfCoastline)
  riskLinesCollection = rbind(riskLinesCollection,
                              riskZone)
  
  # Expand risk zone 2km out to sea.
  riskZone = st_buffer(riskZone, 2000)
  
  # Also tag it background / ambient coastal risk level..
  riskCollection = rbind(riskCollection,
                         cbind(riskZone,maxPathogenSeverityLevel=1)
  )
  
}

# Now deal with isolated ambient stations in low coverage stretches of coastline.
remainingSites=unlist(noProbList[lengths(noProbList)==1])
removeStns=match(alreadyAssigned,remainingSites)
removeStns=removeStns[!is.na(removeStns)]
remainingSites=remainingSites[-removeStns]

for(i in remainingSites)
{
  lIndex=unlist(i)
  
  # Cut out this section of coastline upto and including the next ambient or out of coverage monitoring site at
  # either end.
  # Include adjacent monitoring / end of coverage sites as this is an isolated site.
  
  coastStart=tCoastalStations[(lIndex-1),]$nearestPointsIndices
  coastStop=tCoastalStations[(lIndex+1),]$nearestPointsIndices
  
  riskZone = st_sfc(st_linestring(st_coordinates(irlVertices[coastStart:coastStop,])), crs = 29903)
  riskZone <- st_sf(geometry = riskZone)
  
  # Now record the line of risk along the coast, we will use this to get the surrounding
  # cattle densities etc., later
  # Also tag it with a name of the station and LocationId
  riskZone$Name=tCoastalStations[lIndex,]$Name
  riskZone$maxPathogenSeverityLevel=1
  riskZone$LocationId=list(tCoastalStations[(lIndex-1):(lIndex+1),]$LocationId)
  riskZone$distanceToStartOfCoastline=list(tCoastalStations[(lIndex-1):(lIndex+1),]$distanceToStartOfCoastline)
  riskLinesCollection = rbind(riskLinesCollection,
                              riskZone)
  
  # Expand risk zone 2km out to sea.
  riskZone = st_buffer(riskZone, 2000)
  
  # Also tag it background / ambient coastal risk level..
  riskCollection = rbind(riskCollection,
                         cbind(riskZone,maxPathogenSeverityLevel=1)
  )
}

riskCollectionSf = st_as_sf(data.frame(riskCollection),
                            crs = 29903)

# Now cut out the mainland section from the risk collection to leave us with the coastal waters.
# Do it in 2 steps, Achill Island first then the rest of the country.
riskCollectionSf = st_difference(riskCollectionSf,st_cast(achill,"POLYGON"))
riskCollectionSf = st_difference(riskCollectionSf,irlPoly)


riskCollectionSf$maxPathogenSeverityLevel = unlist(riskCollectionSf$maxPathogenSeverityLevel)
# Sort by severity level.
riskCollectionSf = riskCollectionSf[order(riskCollectionSf$maxPathogenSeverityLevel),]


#####Now plot.

bathingLevelsIconsFiles = c("ICONS/blueBpin.png",
                            "ICONS/greenBpin.png",
                            "ICONS/orangeBpin.png",
                            "ICONS/redBpin.png",
                            "ICONS/purpleBpin.png")

bathingLevelsIcons = tmap_icons(bathingLevelsIconsFiles)

amrSeverityPalett = c("lightblue", "yellow", "orange", "red")

amrSeverityBreaks = c(1, 2, 3, 4, 5)

bbox_new <- st_bbox(repIrl)

xrange <- bbox_new$xmax - bbox_new$xmin # range of x values
yrange <- bbox_new$ymax - bbox_new$ymin # range of y values

bbox_new[1] <- bbox_new[1] - (.1 * xrange) # xmin - left
bbox_new[3] <- bbox_new[3] + (.13 * xrange) # xmax - right
bbox_new[2] <- bbox_new[2] - (.1 * yrange) # ymin - bottom
bbox_new[4] <- bbox_new[4] + (.1 * yrange) # ymax - top


# Setup the map iof Republic of Ireland as no monitoring coverage.
# Regions for which we have coverage will be overlayed with subsequent layers.
tm_shape(irlMainland, bbox=bbox_new) + tm_lines(col = "grey", lwd=1) +
  tm_shape(NorthernIrl) + tm_borders(col="grey", lty="dotted") +
  tm_shape(riskCollectionSf) +
  tm_fill("maxPathogenSeverityLevel", 
          breaks = amrSeverityBreaks,
          style = "fixed",
          textNA = "No data",
          colorNA = "grey", 
          palette = amrSeverityPalett) +
  # For this plot we move the name labels above the EP icons for ease of viewing.
  tm_shape(tCoastalStations[grepl("^North|Balbriggan|Stradbally|Wicklow Harbour|Lady's Bay", tCoastalStations$Name),])+ 
  tm_text("Name", xmod = -2, ymod = -0.1, size=0.5) +
  # tm_shape(tCoastalStations[grepl("Balbriggan|Stradbally|Wicklow Harbour|Lady's Bay", tCoastalStations$Name),]) +
  tm_symbols(shape = "pathogenSeverityLevel", legend.shape.show = FALSE,
             shapes = tmap_icons(bathingLevelsIconsFiles[3:4]),
             size = 1,
             border.col = NULL)
```

### Stage coastal WWTP data

Consolidate WWTP data. Link effluent discharge points with treatment type and PE (Population Equivalent) data associated with the discharge.

```{r}
wwEp = st_read("DATA/wwtpEmissionPoints.geojson")
# Threshold and buffer sizes we use in this chunk.
# Over subscription threshold indicates how oversubscribed (number of people) you need to be to pose a significant threat to discharge water bodies.
OversubCriticalThreshold=500
# Get the impact zone around each station.
impactZoneBufferMetres = 2000

# Add the Agglomeration pop. equivalent of the associated WWTP to each emission point entry.
# Create a RegCD field (strip the version number from RegNo)
wwEp$RegCD = sub("-.*","",wwEp$RegNo)
wwEp$wwtpAgglomerationPE = wwtpFacilities$AgglomerationPE[match(wwEp$RegCD,wwtpFacilities$RegCD)]
wwEp$wwtpPlantDesignPE = wwtpFacilities$PlantDesignPE[match(wwEp$RegCD,wwtpFacilities$RegCD)]

# Add a WWTP "oversubscribed" attribute to each EP, with a value of zero if the WWTP associated with that EP
# is not oversubscribed, or the population size by which it exceeds its design capacity if it is.
wwEp$wwtpOverSubscribed = wwEp$wwtpAgglomerationPE - wwEp$wwtpPlantDesignPE
wwEp$wwtpOverSubscribed[wwEp$wwtpOverSubscribed<0] = 0;

# Get the total number of emission points associated with each WWTP license.
delme = wwEp %>% 
  group_by(RegCD) %>% 
  summarise(totalNumberOfEps = n())

# Add them to the emission points dataframe
# ie., each EP entry is one of a total of totalNumberOfEps that are associated with a given (wwtpFacilities$ActiveLicenceNumber) WWTP
wwEp$totalNumberOfEps = delme$totalNumberOfEps[match(wwEp$RegCD,delme$RegCD)]
delme=NULL

# Add an "oversubscribed" attribute to each EP, which is the total oversubscribed PE of the WWTP associated with the EP, divided by the number of EPs that that WWTP has..
wwEp$epOverSubscribed = wwEp$wwtpOverSubscribed/wwEp$totalNumberOfEps

# Oversubscribed impact is described as the oversubscribed population divided by the number of emission points.

# Add the type of treatment applied by the WWTP associated with this EP
wwEp$TreatmentType = wwtpFacilities$TreatmentType[match(wwEp$RegCD,wwtpFacilities$RegCD)]

# Convert to the same projection
wwEp = st_transform(wwEp, crs=29903)

# Get all EPs (WWTP emission points) of interest.
# Subset the list of all emission points to only contain the emission points within the
# impact zone area of each coastal region of interest.
wwEpSubset=unique(unlist(st_contains(st_buffer(riskLinesCollection,impactZoneBufferMetres), wwEp)))
wwEpBathingLoc = wwEp[wwEpSubset,]

# Remove any emission points that come from WWTPs whose license has been revoked or withdrawn.
# NB: *I am assuming that WWTPs that have had their license revoked or withdrawn are no longer in operation and have no env. AMR impact*.
wwEpBathingLoc$LicenceStatusType = wwtpFacilities$LicenceStatusType[match(wwEpBathingLoc$RegCD,wwtpFacilities$RegCD)]

# Removed NAs from small WWTPs (<500PE) records in the wwtpFacilities table.
wwEpBathingLoc=wwEpBathingLoc[!is.na(wwEpBathingLoc$LicenceStatusType),]

# Remove non avtive licences
wwEpBathingLoc = wwEpBathingLoc[!(wwEpBathingLoc$LicenceStatusType == "Revoked" |
                                    wwEpBathingLoc$LicenceStatusType == "Withdrawn" |
                                    wwEpBathingLoc$LicenceStatusType == "Replaced"),]

# Add an impact level field
wwEpBathingLoc$wwtpAmrImpactLevel = NA

wwEpBathingLoc[wwEpBathingLoc$wwtpOverSubscribed>OversubCriticalThreshold,]$wwtpAmrImpactLevel = "Oversubscribed";
wwEpBathingLoc[grepl("^0", wwEpBathingLoc$TreatmentType),]$wwtpAmrImpactLevel = "No treatment";
wwEpBathingLoc[grepl("^1 ", wwEpBathingLoc$TreatmentType),]$wwtpAmrImpactLevel = "Primary treatment only";

# Remove the better performing WWTPs, we're only interested in the ones of concern for now.
wwEpBathingLoc=wwEpBathingLoc[!is.na(wwEpBathingLoc$wwtpAmrImpactLevel),]

# Note: It is important to remember that:
# For WWTPS with no treatment or primary treatment only, the biggest env. impact form a pathogen perspective is the discharge from the primary emission point(s).
# For oversubscribed WWTPs the biggest env. impact form a pathogen perspective is during times of high load, or also, in the case of combined storm sewer overflows, during heavy rain events.

```
### Pathogen source attribution and AMR weighting

Attribute coastal segment pathogen levels to likely sources, and apply source-specific AMR weighting factors to estimate coastal environmental AMR risk.

```{r, fig.height = 8, fig.width = 7.5, fig.align = "left"}

# Function Name: getEdTotalNumbersInImpactZone
# Description:
# Calculates the total number of units (e.g., cattle, septic tanks, etc.) located within the defined impact zone based on the relevant CSO Electoral Division dataset.
# 
# Parameters:
# 
# sqKmDensity - The column name in the CSO dataset (e.g., totalCattlePerKm) representing the density of units per square kilometre.
# 
# impactZone - The spatial area surrounding the bathing site that influences water quality.
# 
# sfDataSet - The CSO spatial dataset containing density and geographic information for Electoral Divisions.
getEdTotalNumbersInImpactZone = function(sqKmDensity, impactZone, sfDataSet)
{
  intersection_sf <- st_intersection(sfDataSet, impactZone)
  
  # Reset the intersection areas for this subset.
  # Remember in most cases, after it has been intersected with the impact zone we will only
  # be left with a portion of the electoral district so the area must be updated.
  intersection_sf$area = st_area(intersection_sf)
  
  # Assume and electoral divisions which do no have any cattle/septic tank numbers available
  # do not have any cattle
  densityVector = data.frame(intersection_sf)[,sqKmDensity]
  densityVector[is.na(densityVector)] = 0
  
  # Rem. to convert area to km^2
  totalNumbersInImpactZone = as.vector(sum(densityVector * intersection_sf$area/10^6))
  
  return(totalNumbersInImpactZone)
}

# Function Name: getCoastalSegMeanMpn
# Description: Pathogen Loading Estimation

# Pathogen loading for each coastal segment are estimated using a numerical integration approach
# based on 5 year median monitoring data collected at discrete sites along the coastline.
# Pathogen concentrations between monitoring sites are linearly interpolated to generate a
# continuous concentration profile along each segment. The area under this concentration-distance curve
# are then calculated using the trapezoidal rule to provide an estimate of total pathogen loading.
# The mean level for the coastal segment is calculated by dividing by the coastal segment length.
# This approach assumes linear variation in concentration between sites and provides a
# consistent basis for comparing pathogen loading among coastal segments and accounts
# for the relative position of each monitoring site within the segment.


getCoastalSegMeanMpn = function(coastalSegDf, pathogen)
{
  coastalSegmentMeanPathogenMpn=c()
  
  for(i in 1:nrow(coastalSegDf))
  {
    # Read the median ecoli levels for the monitoring sites in this coastline segment.
    pathLevels=unname(unlist(bqSummary[match(unlist(coastalSegDf$LocationId[i]), bqSummary$LocationId),pathogen]))
    
    # Any "end of coverage" monitoring site markers will not have an eColi levels associated with them
    # so set these to 0. TODO!!!! update this with a better estimate of the ambient baseline levels
    pathLevels[is.na(pathLevels)] = 0
    # Get coastal segment monitoring sites / markers and work out its length.
    segment = unlist(coastalSegDf$distanceToStartOfCoastline[i])
    # Segment length = distance from first to last monitoring site / end of coverage site marker.
    coastalSegmentLength = segment[length(segment)] - segment[1]
    coastalSegmentMeanPathogenMpn = rbind(coastalSegmentMeanPathogenMpn,
                                          trapz(unlist(coastalSegDf$distanceToStartOfCoastline[i]),
                                                pathLevels)/coastalSegmentLength)
  }
  return(coastalSegmentMeanPathogenMpn)
}


# Function Name: epaWwtpRawEffluentPEimpact
# Description: Estimates the PE raw sewage discharge equivalent from overloaded or primary treatment only WWTPs or WWTsP not applying any treatment (raw sewage) within the impact zone provide ("buffer" arg).
epaWwtpRawEffluentPEimpact = function(wwEpData, buffer)
{
  
  # Get the emission points (EPs) within the impact zone.
  wwEpDataIdx=unique(unlist(st_contains(buffer, wwEpData)))
  wwEpDataSubset=wwEpData[wwEpDataIdx,]
  
  # Reset out running total.
  rawEffluentPathogenLoadingPE = 0
  
  # Estimate with raw effluent impact from overflow through oversubscribed UWWTPS emission points.
  # Overflow estimates are based on Ringsend UWWTP
  # D0034-02
  # "between 2017 and 2020, 8.9 million cubic metres (8900 million litres) of untreated sewage and storm waters built up after periods of high rainfall and were discharged into Dublin Bay – primarily from overflow tanks at the Ringsend Wastewater Treatment Plant."
  # https://www.irishtimes.com/news/environment/over-one-fifth-of-dublin-bay-water-users-become-ill-pressure-group-claims-1.4536227#:~:text=SOS%20Dublin%20Bay%20says%20it%20has%20analysed,waters%20built%20up%20after%20periods%20of%20high
  
  # Daily raw discharge from ringsend, in litres
  # > (8.9 * 10^9) / (365*3)
  # [1] 8127854
  
  # Ringsend oversubscribed level
  # wwtpOverSubscribed = 511498
  
  # Daily raw discharge per oversubscribed PE in liters per day
  # > 8127854/511498
  # [1] 15.89029
  
  # One population equivalent (PE) typically creates
  # 200 liters of waste water per day.
  # Therefore the percentage of typical PE raw discharge in liters
  # > 15.89029/200
  # [1] 0.07945145
  # Or about 8% of one oversubscribed PE ends up as raw discharge.
  # That is the figure we use in this estimation.
  
  rawEffluentPathogenLoadingPE = rawEffluentPathogenLoadingPE + 0.08 * sum(wwEpDataSubset$epOverSubscribed)
  
  # Now deal with number of primary (ie., main) emission points within the impact zone
  # from WWTPs that apply no sewerage treatment
  
  # Primary emission points
  wwEpDataSubset = wwEpDataSubset[wwEpDataSubset$EmissionType=="Primary",]
  # Remove any duplicates on the same license
  wwEpDataSubset = wwEpDataSubset[!duplicated(wwEpDataSubset$RegNo),]
  
  
  rawEffluentPathogenLoadingPE = rawEffluentPathogenLoadingPE + sum(wwEpDataSubset[wwEpDataSubset$TreatmentType=="0 - No Treatment",]$wwtpAgglomerationPE)
  
  # Now record number of primary emission points within the impact zone
  # from WWTPs that apply primary treatment only
  
  primaryOnlyEffluentPathogenLoadingPE = sum(wwEpDataSubset[wwEpDataSubset$TreatmentType=="1 - Primary Treatment",]$wwtpAgglomerationPE)
  
  # Convert this to an equivalent raw effluent load and add it to our total.
  # Primary treatment removes aprox 30% of e.coli
  # Ref. Raboni et al., DOI 10.1007/s11270-016-3157-8
  # "Assessment of the Fate of Escherichia coli in Different Stages of Wastewater Treatment Plants"
  #
  # Use this approximation to adjust our total raw effluent discharge within the impact zone
  
  rawEffluentPathogenLoadingPE = (0.7 * primaryOnlyEffluentPathogenLoadingPE) + rawEffluentPathogenLoadingPE
  
  return(rawEffluentPathogenLoadingPE)
}

# Set up national data on environmental AMR drivers.

# Intensive agriculture:
# Get geospatial information on poultry and pig farms
# Poultry
poultryFarms=ielFacilities[grepl("^6\\.1 ", ielFacilities$LemaMajorClassOfActivity,),]
# Pigs
pigFarms=ielFacilities[grepl("^6\\.2 ", ielFacilities$LemaMajorClassOfActivity,),]

# Now convert national geodatasets on intensive agriculture, cattle and septic tanks to the same projection
edSfAVA42 = st_transform(edSfAVA42, crs=29903)
edSfSAP2022T6T7ED_2022 = st_transform(edSfSAP2022T6T7ED_2022, crs=29903)
pigFarms = st_transform(pigFarms, crs=29903)
poultryFarms = st_transform(poultryFarms, crs=29903)

# Initialse totals
numCattleInImpactZone = c()
numSepticTankInImpactZone = c()
impactZoneArea = c()
coastalSegmentLength = c()
humanRawEffluentPE = c()
numPigFarms = c()
numPoultryFarms = c()
riskLinesCollection$pigFarmsList=NA
riskLinesCollection$poultryFarmsList=NA

# Go through each coastal segment and estimate pathogen contributions from each of the main environmental AMR drivers.
for(i in 1:nrow(riskLinesCollection))
{
  tmpBuff = st_buffer(riskLinesCollection[i,],impactZoneBufferMetres)
  tmpBuffArea = st_area(tmpBuff)
  numCattleInImpactZone = c(numCattleInImpactZone,
                            getEdTotalNumbersInImpactZone("totalCattlePerKm",
                                                          tmpBuff, edSfAVA42))
  numSepticTankInImpactZone = c(numSepticTankInImpactZone,
                                getEdTotalNumbersInImpactZone("individualSepticTankPerKm",
                                                              tmpBuff,
                                                              edSfSAP2022T6T7ED_2022))
  
  humanRawEffluentPE =  c(humanRawEffluentPE,
                          epaWwtpRawEffluentPEimpact(wwEpBathingLoc, tmpBuff))
  
  numPigFarms =  c(numPigFarms,length(unique(unlist(st_contains(tmpBuff, pigFarms)))))
  numPoultryFarms =  c(numPoultryFarms,length(unique(unlist(st_contains(tmpBuff, poultryFarms)))))
  
  riskLinesCollection$pigFarmsList[i]=list(pigFarms$RegCD[unique(unlist(st_contains(tmpBuff, pigFarms)))])
  riskLinesCollection$poultryFarmsList[i]=list(poultryFarms$RegCD[unique(unlist(st_contains(tmpBuff, poultryFarms)))])
  
  # Also note coastline length and impact zone area, as these will be useful below.
  impactZoneArea = c(impactZoneArea,tmpBuffArea)
  
  # Get coastal segment monitoring sites / markers and work out its length.
  segment = unlist(riskLinesCollection$distanceToStartOfCoastline[i])
  # Segment length = distance from first to last monitoring site / end of coverage site marker.
  coastalSegmentLength = c(coastalSegmentLength,segment[length(segment)] - segment[1])
}

# Update update each coastal segment within the collection with the level of each
# environmental AMR driver within each impact zone.
riskLinesCollection$impactZoneArea = impactZoneArea
#riskLinesCollection$humanRawEffluentPE = coastalSegmentLength
riskLinesCollection$numCattleInImpactZone = numCattleInImpactZone
riskLinesCollection$numSepticTankInImpactZone = numSepticTankInImpactZone
riskLinesCollection$numPigFarms = numPigFarms
riskLinesCollection$numPoultryFarms = numPoultryFarms
riskLinesCollection$humanRawEffluentPE = humanRawEffluentPE

# From empirical, BWQ monitoring data, estimate the MPN loadings per metre of each coastal segment.
# We use linear interpolation (join the dots) to fill in the missing data between monitoring sites within the coastal segment
# and then the trapezoid rule to calculate the area under the coastline pathogen profile and use this as pathogen loading
# estimate for the coastal segment, diving by the coastline length to get an average MPN for that coastal segment.
# The simplest way is to use trapz from the pracma package to use the trapezoidal method for numerical integration.
ecoliEmpiricalMpnCoastalSegmentMean=getCoastalSegMeanMpn(riskLinesCollection, "EcoliResult")
riskLinesCollection$ecoliEmpiricalMpnCoastalSegmentMean = ecoliEmpiricalMpnCoastalSegmentMean
enterococciEmpiricalMpnCoastalSegmentMean=getCoastalSegMeanMpn(riskLinesCollection, "EnterococciResult")
riskLinesCollection$enterococciEmpiricalMpnCoastalSegmentMean=enterococciEmpiricalMpnCoastalSegmentMean


# Convert all AMR drivers impact on coastal waters to common scale;
# Work out MPN footprints for each of the AMR drivers in the impact zone
# MPN footprints are listed per metre of coastal segment.

# The cattle MPN equivalent was defined as the effective contribution of an individual cow to the coastal pathogen load. It was calculated as the product of two scaling factors: (1) the fraction of daily pathogen loading generated per head of cattle relative to that produced per human, and (2) the fraction of viable pathogen loading transferred to coastal waters via runoff.

# WEIGHTED OVERLAY ANALYSIS OF AMR PREVALENCE IN BATHING WATERS

# STEP 1: Reclassify Data: Convert to common data type and scale
# 
# We first convert the scale of each drivers within a coastal segment’s impact zone into their estimated pathogen contributions to that coastal segment.

# Reclassify empirical E.coli data first.

# Daily loading per head of cattle = 47.14 x 10^14 MPN
# Daily loading per PE person = 200 x 6.2 X 10^6 x 10^6 = 1240 x 10^12 = 12.4 x 10^14 MPN
# Runoff factor (fraction of viable pathogens that reach coastal waters from cattle effluent)

# Sewage, runoff_factor ~= 1
# Daily loading per PE person (aprox 200l wastewater / day) = 200 x 6.2 X 10^6 x 10^6 = 1240 x 10^12 = 12.4 x 10^14 MPN / liter /day
riskLinesCollection$humanEcoliMPNfootprint = humanRawEffluentPE * (12.4 * 10^14) / coastalSegmentLength


# Cattle loading, runoff_factor ~= 0.05
# Daily loading per head of cattle = 47.14 x 10^6 x 10^8 x runoff_factor = 47.14 x 10^14 MPN x 0.05
riskLinesCollection$cattleEcoliMPNfootprint = numCattleInImpactZone * (47.14 * 10^14) * 0.05 / coastalSegmentLength

# Septic tanks
# 2.1 x 10^6 MPN per 100ml per 4 PE tank.. 
# Multiply your 200l/day below by 4 for a 4PE tank = 800l effluent per day
# Daily loading = (2.1 x 10^6 x 10^6) per liter  X  800 liters per day = 1680 x 10^12 = 16.8 x 10^14 MPN / liter /day
# One study estimates that upto 15% of pathogens from septic tanks make it into the groundwater in Ireland.
# "this methodology suggests that, on a net national average, 15.5% of pathogens or phosphorus and 16.0% of nitrogen from DWWTS make it through the unsaturated subsoil to groundwater (assuming that there is a uniform density of DWWTSs across the country)."
# https://www.mdpi.com/2071-1050/6/3/1623
# The percentage of viable pathogen making it to the seq via diversw runoff mechanism is likely much less, lets say 5%
# runoff_factor ~= 0.05

# Adjust daily septic tank loading at coast = 16.8 x 10^14 x 0.05
riskLinesCollection$septicTankEcoliMPNfootprint = numSepticTankInImpactZone * (16.8 * 10^14 * 0.05) / coastalSegmentLength

# Bathing water pathogen levels within a coastal segment depend not only on pathogen inputs
# from sources within its impact zone, but also on the capacity of adjacent coastal waters
# to disperse and dilute pathogen loads.
# We know that these estimates do not account for do not account for marine dispersion or dilution. 
# To correct for this we scale these estimates with empirical pathogen data from historical bathing water surveillance.


# Calibrate the estimated pathogen levels for each coastal segment with
# empirical data from coastal bathing water surveillance.

riskLinesCollection$totalEcoliMPNfootprint = riskLinesCollection$septicTankEcoliMPNfootprint + riskLinesCollection$cattleEcoliMPNfootprint + riskLinesCollection$humanEcoliMPNfootprint


# Quantify the contributions of all relevant AMR sources to E. coli MPN levels as observed in surveillance datasets.
riskLinesCollection$cattleEcoliMPN = (riskLinesCollection$cattleEcoliMPNfootprint/riskLinesCollection$totalEcoliMPNfootprint)*riskLinesCollection$ecoliEmpiricalMpnCoastalSegmentMean
riskLinesCollection$humanEcoliMPN = (riskLinesCollection$humanEcoliMPNfootprint/riskLinesCollection$totalEcoliMPNfootprint)*riskLinesCollection$ecoliEmpiricalMpnCoastalSegmentMean
riskLinesCollection$septicTankEcoliMPN = (riskLinesCollection$septicTankEcoliMPNfootprint/riskLinesCollection$totalEcoliMPNfootprint)*riskLinesCollection$ecoliEmpiricalMpnCoastalSegmentMean

# Weight each of these contributions by the expected prevalence of AMR in e.coli from each source.
# These are rough estimates from the literature, ie.,
# https://www.sciencedirect.com/science/article/abs/pii/S0048969723008100
# https://www.science.org/doi/10.1126/sciadv.adt8073
# https://www.cidrap.umn.edu/antimicrobial-stewardship/livestock-manure-contains-antibiotic-resistance-genes-posing-health

riskLinesCollection$cattleEcoliAmrMPN = 0.5 * riskLinesCollection$cattleEcoliMPN
riskLinesCollection$humanEcoliAmrMPN = 0.9 * riskLinesCollection$humanEcoliMPN
riskLinesCollection$septicTankEcoliAmrMPN = 0.85 * riskLinesCollection$septicTankEcoliMPN

riskLinesCollection$ecoliAmrMpnCoastalSegmentMean = riskLinesCollection$cattleEcoliAmrMPN + riskLinesCollection$humanEcoliAmrMPN + riskLinesCollection$septicTankEcoliAmrMPN

# Now reclassify empirical Enterococci data.

# Daily loading per head of cattle = 47.14 x 10^14 MPN
# Daily loading per PE person = 200 x 0.6 X 10^6 x 10^6 = 1240 x 10^12 = 12.4 x 10^14 MPN
# Runoff factor (fraction of viable pathogens that reach coastal waters from cattle effluent)

# Sewage, runoff_factor ~= 1
# Daily loading per PE person (aprox 200l wastewater / day) = 200 x 10^6 x 0.6  x 10^6 = 1240 x 10^12 = 1.2 x 10^14 MPN / liter /day
# Human sewage runoff factor is 1 as it is pumped directly into the sea.
riskLinesCollection$humanEnterococciMPNfootprint = humanRawEffluentPE * (1.2 * 10^14) / coastalSegmentLength

# Cattle loading, runoff_factor ~= 0.05
# Daily loading per head of cattle = 47.14 x 10^6 x 10^7 x runoff_factor = 47.14 x 10^14 MPN x 0.05
riskLinesCollection$cattleEnterococciMPNfootprint = numCattleInImpactZone * (47.14 * 10^13) * 0.05 / coastalSegmentLength

# Septic tanks
# 0.2 x 10^6 MPN per 100ml per 4 PE tank.. 
# Multiply your 200l/day below by 4 for a 4PE tank = 800l effluent per day
# Daily loading = (0.2 x 10^6 x 10^6) per liter  X  800 liters per day = 160 x 10^12 = 1.6 x 10^14 MPN / liter /day
# One study estimates that upto 15% of pathogens from septic tanks make it into the groundwater in Ireland.
# "this methodology suggests that, on a net national average, 15.5% of pathogens or phosphorus and 16.0% of nitrogen from DWWTS make it through the unsaturated subsoil to groundwater (assuming that there is a uniform density of DWWTSs across the country)."
# https://www.mdpi.com/2071-1050/6/3/1623
# The percentage of viable pathogen making it to the seq via diversw runoff mechanism is likely much less, lets say 5%
# runoff_factor ~= 0.05

# Adjust daily septic tank loading at coast = 1.6 x 10^14 x 0.05
riskLinesCollection$septicTankEnterococciMPNfootprint = numSepticTankInImpactZone * (1.6 * 10^14 * 0.05) / coastalSegmentLength

# We now scale these estimates with empirical pathogen data from historical bathing water surveillance.

# Bathing water pathogen levels within a coastal segment depend not only on pathogen inputs
# from sources within its impact zone, but also on the capacity of adjacent coastal waters
# to disperse and dilute these loadings.

# To account for this we calibrate the estimated pathogen levels for each coastal segment with
# empirical data from coastal bathing water surveillance.

riskLinesCollection$totalEnterococciMPNfootprint = riskLinesCollection$septicTankEnterococciMPNfootprint + riskLinesCollection$cattleEnterococciMPNfootprint + riskLinesCollection$humanEnterococciMPNfootprint


# Quantify the contributions of all relevant AMR sources to enterococci NPM levels as observed in surveillance datasets.
riskLinesCollection$cattleEnterococciMPN = (riskLinesCollection$cattleEnterococciMPNfootprint/riskLinesCollection$totalEnterococciMPNfootprint)*riskLinesCollection$enterococciEmpiricalMpnCoastalSegmentMean
riskLinesCollection$humanEnterococciMPN = (riskLinesCollection$humanEnterococciMPNfootprint/riskLinesCollection$totalEnterococciMPNfootprint)*riskLinesCollection$enterococciEmpiricalMpnCoastalSegmentMean
riskLinesCollection$septicTankEnterococciMPN = (riskLinesCollection$septicTankEnterococciMPNfootprint/riskLinesCollection$totalEnterococciMPNfootprint)*riskLinesCollection$enterococciEmpiricalMpnCoastalSegmentMean

# Weight each of these contributions by the expected prevalence of AMR in enterococci from each source.
# These are very rough estimates from the literature
# https://www.sciencedirect.com/science/article/abs/pii/S0048969723008100
# https://www.cidrap.umn.edu/antimicrobial-stewardship/livestock-manure-contains-antibiotic-resistance-genes-posing-health

# Heuristic weights for prevalence of AMR in cattle feaces vs raw sewage vs septic tanksetc.
riskLinesCollection$cattleEnterococciAmrMPN = 0.5 * riskLinesCollection$cattleEnterococciMPN
riskLinesCollection$humanEnterococciAmrMPN = 0.9 * riskLinesCollection$humanEnterococciMPN
riskLinesCollection$septicTankEnterococciAmrMPN = 0.85 * riskLinesCollection$septicTankEnterococciMPN

riskLinesCollection$enterococciAmrMpnCoastalSegmentMean = riskLinesCollection$cattleEnterococciAmrMPN + riskLinesCollection$humanEnterococciAmrMPN + riskLinesCollection$septicTankEnterococciAmrMPN

# AMR risk, Enterococci vs E.coli ?
#
# Enterococci are more resilient than E. coli in environmental settings, allowing
# them to persist and carry ARGs for longer periods. They are also more resilient in salwater.
# However some studies indicate there may be slightly higher levels of resistance
# in E.coli vs enterococci (PMC12561136, PMID: 25252346).
# I could not decide how to weight one vs the other.
# So in the end they both got a weighting of 1.
# Update these and other weightings accordingly as more detailed information becomes available.

enterococciAmrWeight = 1
ecoliAmrWeight = 1

# Get the total prevalence estimate
riskLinesCollection$amrCoastalPrevalence = (ecoliAmrWeight * riskLinesCollection$ecoliAmrMpnCoastalSegmentMean) + (enterococciAmrWeight * riskLinesCollection$enterococciAmrMpnCoastalSegmentMean)

elevatedRisk=riskLinesCollection


# Extend a 2km buffer out from each coastal segment.
# We fill this buffer according to parameter of interest below
# (ie., pathogen level/ env. AMR risk etc)
# Create the buffer then cut out the mainland section from the
# risk collection to leave us with the coastal waters.
# Do it in 2 steps, Achill Island first then the rest of the country.
tmpSf = st_difference(st_buffer(elevatedRisk, 4000),st_cast(achill,"POLYGON"))
tmpSf = st_difference(tmpSf,irlPoly)

# Sort by severity level.
tmpSf = tmpSf[order(tmpSf$maxPathogenSeverityLevel),]

#tm_shape(st_rasterize(tmpSf)) +
#  tm_raster("cattleEcoliMPNfootprint", palette = "Reds")

################################
# Enlarge bounding box to accommodate legend.

bbox_new <- st_bbox(repIrl)

bbox_new[1] <- bbox_new[1] * 1.13 # xmin - left
bbox_new[3] <- bbox_new[3] * 1 # xmax - right
bbox_new[2] <- bbox_new[2] * 1 # ymin - bottom
bbox_new[4] <- bbox_new[4] * 1.015 # ymax - top


# AMR drivers rasters

tm_shape(repIrl, bbox = bbox_new) + tm_lines(lwd = 0.1) +
  tm_shape(st_rasterize(tmpSf[order(tmpSf$cattleEcoliMPNfootprint),])) +
  tm_raster("cattleEcoliMPNfootprint", style = "cont", palette = "matplotlib.yl_or_rd", labels = c(" 0", " 2 × 10 ⁴", " 4 × 10 ⁴", " 6 × 10 ⁴", " 8 × 10 ⁴", " 10 × 10 ⁴"), title = "") + tm_layout(
    title = "A: Modeled contribution of cattle\nto coastal escherichia coli (MPN)",
    title.size = 2,
    legend.title.size = 2,
    legend.position = c("left", "bottom"),
    legend.bg.alpha = 0,
    legend.text.size = 1,
    legend.frame.color = "#00000000",
    bg.color = "#00000000",
    frame=F)

tm_shape(repIrl, bbox = bbox_new) + tm_lines(lwd = 0.1) +
  tm_shape(st_rasterize(tmpSf[order(tmpSf$septicTankEcoliMPNfootprint),])) +
  tm_raster("septicTankEcoliMPNfootprint", style = "cont", palette = "matplotlib.yl_or_rd", labels = c("0.5 × 10 ³","1 × 10 ³","1.5 × 10 ³","2 × 10 ³","2.5 × 10 ³","3 × 10 ³","3.5 × 10 ³"), title = "") +
  tm_layout(
    title = "B: Modeled contribution of septic tanks\nto coastal escherichia coli (MPN)",
    title.size = 2,
    legend.title.size = 2,
    legend.position = c("left", "bottom"),
    legend.bg.alpha = 0,
    legend.text.size = 1,
    legend.frame.color = "#00000000",
    bg.color = "#00000000",
    frame=F)


tm_shape(repIrl, bbox = bbox_new) + tm_lines(lwd = 0.1) +
  tm_shape(st_rasterize(tmpSf[order(tmpSf$humanEcoliMPNfootprint),])) +
  tm_raster("humanEcoliMPNfootprint", style = "cont", palette = "matplotlib.yl_or_rd", labels = c(" 0", " 5 × 10 ⁴", " 10 × 10 ⁴", " 15 × 10 ⁴", " 20 × 10 ⁴", " 25 × 10 ⁴", " 30 × 10 ⁴"), title = "") +
  tm_layout(
    title = "C: Modeled contribution of sewage\nto coastal escherichia coli (MPN)",
    title.size = 2,
    legend.title.size = 2,
    legend.position = c("left", "bottom"),
    legend.bg.alpha = 0,
    legend.text.size = 1,
    legend.frame.color = "#00000000",
    bg.color = "#00000000",
    frame=F)


tm_shape(repIrl, bbox = bbox_new) + tm_lines(lwd = 0.1) +
  tm_shape(st_rasterize(tmpSf[order(tmpSf$ecoliEmpiricalMpnCoastalSegmentMean),])) +
  tm_raster("ecoliEmpiricalMpnCoastalSegmentMean", style = "cont", palette = "matplotlib.yl_or_rd", title = "") +
  tm_layout(
    title = "D: Empirical escherichia coli\nestimates (MPN)",
    title.size = 2,
    legend.title.size = 2,
    legend.position = c("left", "bottom"),
    legend.bg.alpha = 0,
    legend.text.size = 1,
    legend.frame.color = "#00000000",
    bg.color = "#00000000",
    frame=F)

tm_shape(repIrl, bbox = bbox_new) + tm_lines(lwd = 0.1) +
  tm_shape(st_rasterize(tmpSf[order(tmpSf$enterococciEmpiricalMpnCoastalSegmentMean),])) +
  tm_raster("enterococciEmpiricalMpnCoastalSegmentMean", style = "cont", palette = "matplotlib.yl_or_rd", title = "") +
  tm_layout(
    title = "E: Empirical enterococci\nestimates (MPN)",
    title.size = 2,
    legend.title.size = 2,
    legend.position = c("left", "bottom"),
    legend.bg.alpha = 0,
    legend.text.size = 1,
    legend.frame.color = "#00000000",
    bg.color = "#00000000",
    frame=F)

#amrCoastalPrevalence
tm_shape(repIrl, bbox = bbox_new) + tm_lines(lwd = 0.1) +
  tm_shape(st_rasterize(tmpSf[order(tmpSf$amrCoastalPrevalence),])) +
  tm_raster("amrCoastalPrevalence", style = "cont", palette = "matplotlib.yl_or_rd", title = "") +
  tm_layout(
    title = "F: Putative coastal AMR prevalence",
    title.size = 2,
    legend.title.size = 2,
    legend.position = c("left", "bottom"),
    legend.bg.alpha = 0,
    legend.text.size = 1,
    legend.frame.color = "#00000000",
    bg.color = "#00000000",
    frame=F)

# Display intensive agriculture coastal densities.
# The very low densities demonstrates that intensive agriculture does not play a substantive
# role in coastal pathogen levels.
tm_shape(repIrl, bbox = bbox_new) + tm_lines(lwd = 0.1) +
  tm_shape(st_rasterize(tmpSf[order(tmpSf$numPoultryFarms),])) +  tm_raster("numPoultryFarms", style = "cont", palette = "matplotlib.yl_or_rd", title = "") +
  tm_layout(
    title = "Number of poultry farms within impact zone",
    title.size = 2,
    legend.title.size = 2,
    legend.position = c("left", "bottom"),
    legend.bg.alpha = 0,
    legend.text.size = 1,
    legend.frame.color = "#00000000",
    bg.color = "#00000000",
    frame=F)

tm_shape(repIrl, bbox = bbox_new) + tm_lines(lwd = 0.1) +
  tm_shape(st_rasterize(tmpSf[order(tmpSf$numPigFarms),])) +    tm_raster("numPigFarms", style = "cont", palette = "matplotlib.yl_or_rd", title = "") +
  tm_layout(
    title = "Number of pig farms within impact zone",
    title.size = 2,
    legend.title.size = 2,
    legend.position = c("left", "bottom"),
    legend.bg.alpha = 0,
    legend.text.size = 1,
    legend.frame.color = "#00000000",
    bg.color = "#00000000",
    frame=F)

```

### AMR risk assessment case-study: Lough Swilly
```{r, fig.height = 8, fig.width = 7.5, fig.align = "left"}
# Load monitoring stations
tStations = st_as_sf(epaBqXyrsMediansStaged,
                     coords = c("Easting", "Northing"),
                     crs = 29903)


# Buncrana

cSegment = riskLinesCollection[grep("Lady's Bay, Buncrana", riskLinesCollection$Name),]

# There are no relevant industrial emissions licensed facilities nearby
# as evidenced by the following
#ielSubset=unique(unlist(st_intersects(st_buffer(st_zm(cSegment),10000),ielFacilities)))
#ielSubset=ielFacilities[ielSubset,]
#ielSubset

# Get all monitoring sites in this segment
monSites = unlist(cSegment$LocationId)
monSites = monSites[!is.na(monSites)]
monSites=tStations[match(monSites,tStations$LocationId),]

# Get a list of high risk WWTP emission points
wwEpPrimaryBathingLoc=wwEpBathingLoc[!(wwEpBathingLoc$EmissionType!="Primary" & wwEpBathingLoc$wwtpAmrImpactLevel != "Oversubscribed"),]
wwepsLevelsIcons = tmap_icons(c(
  "/home/sully/BIOINFORMATICS/AMR/BATHING_WATERS/ICONS/redAlertIcon.png",
  "/home/sully/BIOINFORMATICS/AMR/BATHING_WATERS/ICONS/yellowAlertIcon.png",
  "/home/sully/BIOINFORMATICS/AMR/BATHING_WATERS/ICONS/purpleAlertIcon.png"
))

# Get a list of high risk WWTP emission points within the impact zone of this coastal segment.
epSubsetIdx=unique(unlist(st_contains(st_buffer(cSegment,impactZoneBufferMetres), wwEpPrimaryBathingLoc)))
epSubset=wwEpPrimaryBathingLoc[epSubsetIdx,]

# Get the WWTP agglomerations associated with WWTP emission points of concern.
wwtpAgglomerations = st_transform(wwtpAgglomerations, crs=29903)
wwtpAgglomSubset = wwtpAgglomerations[match(sub("-.*","",epSubset$RegNo),wwtpAgglomerations$RegCd),]

# Load health care facilities (we intersect this with WWTP agglomerations below)
hiqaNursingHomes=read.table(file='DATA/hiqa_nursing_home_eircodes_and_coordinates.tsv', sep='\t', quote = NULL, header = TRUE)

hiqaNursingHomes = st_as_sf(hiqaNursingHomes,
                            #coords = c("LatitudeDD", "LongatudeDD"),
                            coords = c("LongatudeDD","LatitudeDD"),
                            crs = 4326)
hiqaNursingHomes = st_transform(hiqaNursingHomes, crs=29903)

cHomesSubset = hiqaNursingHomes[unique(unlist(st_contains(wwtpAgglomSubset, hiqaNursingHomes))),]

# Add labels for map
cHomesSubset$Label=paste0(cHomesSubset$Name,", ",cHomesSubset$Capacity," bed")

# Add major rivers/lakes for context.
sfWfdRiverWaterBodies = st_as_sf(wfdRiverWaterBodies,
                                 crs = 4326)
sfWfdRiverWaterBodies = st_transform(sfWfdRiverWaterBodies, crs=29903)


riverSubset=unique(unlist(st_intersects(st_buffer(st_zm(cSegment),30000),sfWfdRiverWaterBodies)))
riverSubset=sfWfdRiverWaterBodies[riverSubset,]
riverSubset=riverSubset[grepl("Crana|Leannan|Swilly", riverSubset$EPA_Name,ignore.case = T),]
riverSubset=st_difference(riverSubset,cSegment)


# Setup a reasonably sized bounding box to give some context.
bbox_new <- st_bbox(st_buffer(monSites[4,], 30000))


xrange <- bbox_new$xmax - bbox_new$xmin # range of x values
yrange <- bbox_new$ymax - bbox_new$ymin # range of y values

bbox_new[1] <- bbox_new[1] - (.1 * xrange) # xmin - left
bbox_new[3] <- bbox_new[3] + (.25 * xrange) # xmax - right
bbox_new[2] <- bbox_new[2] - (.1 * yrange) # ymin - bottom
bbox_new[4] <- bbox_new[4] + (.1 * yrange) # ymax - top

# Set the layout first
tm_layout(legend.outside = TRUE,legend.position = c("right", "top"),
          legend.outside.size = 0.4, 
          legend.text.size = 0.4,
          legend.title.size=0.6) +

# National boundaries.
tm_shape(repIrl, bbox = bbox_new) + tm_lines(lwd = 0.1) + tm_shape(tmpSf[grep("Lady", tmpSf$Name),]) + tm_borders(col="#0000FF", fill="#0099FF11") +
  tm_shape(NorthernIrl) + tm_borders(fill="lightgrey") +
  
  # Cattle densities
  tm_shape(edSfAVA42)+
  tm_fill("totalCattlePerKm",style = "cont", palette = "matplotlib.yl_or_rd", title = "CSO 2020 Animal census, cattle density per km²") + 
  
  # Add aquaculture sites (if any), highlighting potential secondary transmission routes.
  tm_shape(ielAquaculture) + tm_borders(fill="#E8E8E8",
                                        col="#111111", lwd = 0.7) +
  # Buncrana boundary for context
  tm_shape(buncranaBoundary) + tm_fill(col = "#559988FF", fill="#999988FF") +
  
  # Rivers
  tm_shape(riverSubset) + tm_lines(col = "#5599FFFF",lwd=2) +
  
  # Add EP numbers from substandard WWTPs (no or primary treatment only) in the area
  tm_shape(wwEpPrimaryBathingLoc[grepl("^[01] ", wwEpPrimaryBathingLoc$TreatmentType),])+ 
  tm_text("RegNo", xmod = -2, ymod = -0.5, bgcol = "yellow", size=0.2) +
  
  # Finally add the problem WWTP discharge points
  tm_shape(wwEpPrimaryBathingLoc) +
  tm_symbols(shape = "wwtpAmrImpactLevel", size=0.4,
             legend.size.show = FALSE, 
             legend.col.show = FALSE,
             shapes = wwepsLevelsIcons) +
  
  # Bathingwater monitoring sites
  tm_shape(monSites) +
  tm_symbols(shape = "ecoliEnterococciCluster",
             title.shape = "Bathing water quality level:\nE.coli, Enterococci cluster centres",
             shapes = bathingLevelsIcons,
             size = 0.6) +
  tm_text("Name", xmod = 0, ymod = 0.22, size=.7, alpha = 0.3, col = "#89EE8890") +
  
  # Impact zone
  tm_shape(st_buffer(cSegment,impactZoneBufferMetres)) +
  tm_borders(fill="#00000000", col="#00FFAA", lwd = 0.7, lty = "dashed") +
  # Health care facilities within sub-standard UWWTP agglomerations
  tm_shape(cHomesSubset) +tm_symbols(shape=tmap_icons("DATA/careHomeIcon.png"), size=0.13) + tm_text("Label", xmod = 0, ymod = 0.22, size=.64, alpha = 0.4, col = "#89512990") +
  tm_scale_bar(
    breaks = c(0, 2, 4, 6, 8),
    text.size = 0.4
  )


# Driver densities
# Display the AMR driver densities within the impact zone of the lough.
driverDensitiesAroundWaterbody(st_zm(cSegment), impactZoneBufferMetres, "totalCattle", edSfAVA42)
driverDensitiesAroundWaterbody(st_zm(cSegment), impactZoneBufferMetres, "individualSepticTank", edSfSAP2022T6T7ED_2022)

```
Plot all available e.coli monitoring data for lowest water quality bathing locations in Lough Swilly.
```{r, fig.height = 8, fig.width = 7.5, fig.align = "left"}
ecoliWQoverview(epaBathingLocations, bqDataComplete, "Lady's Bay, Buncrana")

```

### AMR risk assessment case-study: Dublin Bay
```{r, fig.height = 8, fig.width = 7.5, fig.align = "left"}

cSegment = riskLinesCollection[grep("Merrion Strand", riskLinesCollection$Name),]

# Get all monitoring sites in this segment

monSites = unlist(cSegment$LocationId)
monSites = monSites[!is.na(monSites)]
monSites=tStations[match(monSites,tStations$LocationId),]

# Get a list of high risk WWTP emission points
wwEpPrimaryBathingLoc=wwEpBathingLoc[!(wwEpBathingLoc$EmissionType!="Primary" & wwEpBathingLoc$wwtpAmrImpactLevel != "Oversubscribed"),]
wwepsLevelsIcons = tmap_icons(c(
  "ICONS/redAlertIcon.png",
  "ICONS/yellowAlertIcon.png",
  "ICONS/purpleAlertIcon.png"
))

# Get a list of high risk WWTP emission points within the impact zone of this coastal segment.
epSubsetIdx=unique(unlist(st_contains(st_buffer(cSegment,impactZoneBufferMetres), wwEpPrimaryBathingLoc)))
epSubset=wwEpPrimaryBathingLoc[epSubsetIdx,]

# And get the list of the agglomerations associated with these emission points.
wwtpAgglomerations = st_transform(wwtpAgglomerations, crs=29903)
wwtpAgglomSubset = wwtpAgglomerations[match(sub("-.*","",unique(epSubset$RegNo)),wwtpAgglomerations$RegCd),]


# Load health care facilities (we intersect this with WWTP agglomerations below)
hiqaNursingHomes=read.table(file='DATA/hiqa_nursing_home_eircodes_and_coordinates.tsv', sep='\t', quote = NULL, header = TRUE)

hiqaNursingHomes = st_as_sf(hiqaNursingHomes,
                            coords = c("LongatudeDD","LatitudeDD"),
                            crs = 4326)
hiqaNursingHomes = st_transform(hiqaNursingHomes, crs=29903)

# Add labels for map
hiqaNursingHomes$Label=paste0(hiqaNursingHomes$Name,", ",hiqaNursingHomes$Capacity," bed")

cHomesSubset = hiqaNursingHomes[unique(unlist(st_contains(wwtpAgglomSubset, hiqaNursingHomes))),]

cHomesSubset$Label=paste0(cHomesSubset$Name,", ",cHomesSubset$Capacity," bed")

# Load OpenStreetMap hospital data
hospitalsSf = st_read("DATA/irelandHospitals.geojson")
ringsendWwtpSf = st_read("DATA/ringsendWwtp.geojson")

# Identify which entries are duplicated between care homes geolocated from HIQA server and hospitals identified from OSM
duplicateHospitals = st_intersects(hospitalsSf, hiqaNursingHomes)
duplicateHospitals = which(lengths(duplicateHospitals) > 0)

# Remove duplicates
hospitalsSf = hospitalsSf[-duplicateHospitals, ]

# Subset to those located within Ringsend treatment plant agglomeration.
hospitalSubset = hospitalsSf[unique(unlist(st_contains(wwtpAgglomSubset, hospitalsSf))),]

# Remove any NA's
hospitalSubset = hospitalSubset[!is.na(hospitalSubset$name),]

# Remove duplicate hospital entries like,
# [30] "Children's Health Ireland at Crumlin"
# [31] "Mater Private Hospital"
hospitalSubset=hospitalSubset[-(30:31),]

# Some facilities listed here (e.g., psychiatric or therapy facilities) do not routinely
# prescribe antibiotics to residents.
#
# An online review was used to estimate which hospitals and residential/long-stay care
# facilities within the Ringsend agglomeration are likely to routinely prescribe antibiotics.
# Estimated bed capacities were also compiled where available.

hospitalBedCapacity = c(258,744,179,NA,47,960,248,589,210,NA,107,150,112,112,176,NA,450,NA,211,127,NA,NA,NA,236,NA,NA,NA,176,165,NA,NA,NA,NA,NA)

# Add this information to the subset
hospitalSubset$hospitalBedCapacity=hospitalBedCapacity

# Remove NAs from the subset
hospitalSubset=hospitalSubset[!is.na(hospitalSubset$hospitalBedCapacity),]

# Add labels for map
hospitalSubset$Label=paste0(hospitalSubset$name,", ",hospitalSubset$hospitalBedCapacity," bed")

# Assume average operating capacity of hospitals in Ireland to determine occupancy on any given day (90%)

# Estimate the current occupancy of hospitals and care homes within the Ringseng WWTP agglomeration.
totalNumPatientsWithinAgglomeration = sum(hospitalSubset$hospitalBedCapacity[!is.na(hospitalSubset$hospitalBedCapacity)]) + sum(cHomesSubset$Capacity[!is.na(cHomesSubset$Capacity)])

# Scale with average operating capacity (90%) of hospitals in Ireland to determine occupancy on any given day
totalNumPatientsWithinAgglomeration = totalNumPatientsWithinAgglomeration * .9

print(paste0("Estimated size of in-patient and care population within the Ringsend agglomeration is ",totalNumPatientsWithinAgglomeration))

# So there is a population equivalent in excess of 10,000 in-patients or care home residents contributing to influent within the Ringsend WWTP agglomeration.


# Prepare main plot...

bbox_new <- st_bbox(st_buffer(cSegment, 12000))

xrange <- bbox_new$xmax - bbox_new$xmin # range of x values
yrange <- bbox_new$ymax - bbox_new$ymin # range of y values

bbox_new[1] <- bbox_new[1] - (.1 * xrange) # xmin - left
bbox_new[3] <- bbox_new[3] + (.25 * xrange) # xmax - right
bbox_new[2] <- bbox_new[2] - (.1 * yrange) # ymin - bottom
bbox_new[4] <- bbox_new[4] + (.1 * yrange) # ymax - top

# Set the layout first
tm_layout(legend.outside = TRUE,legend.position = c("right", "top"),
          legend.outside.size = 0.4, 
          legend.text.size = 0.4,
          legend.title.size=0.6) +
  
  # National boundaries.
  tm_shape(repIrl, bbox = bbox_new) + tm_lines(lwd = 0.1) + tm_shape(tmpSf[grep("Merrion Strand", tmpSf$Name),]) + tm_borders(col="#00AAFF44", fill="#0099FF11") +
  tm_shape(NorthernIrl) + tm_borders(fill="lightgrey") +
  
  # There are very few cattle in the impact zone here so just ignore.
  # Septic tank data is a more informative background in this case.
  tm_shape(edSfSAP2022T6T7ED_2022)+
  tm_fill("individualSepticTankPerKm",style = "cont", palette = "matplotlib.yl_or_rd", title = "Private household 2022 census,\nseptic tank numbers per km²") + 
  
  # Add aquiculture sites (if any), highlighting potential secondary transmission routes.
  tm_shape(ielAquaculture) + tm_borders(fill="#E8E8E8",
                                        col="#111111", lwd = 0.7) +
  
  # Add rivers subset of interest at this site.
  tm_shape(sfWfdRiverWaterBodies[grepl("^09_", sfWfdRiverWaterBodies$Segment_Code,ignore.case = T),]) + tm_lines(col = "#0000FF70", fill="#0000FF70",lwd=2) +
  
  # Add EP numbers from substandard WWTPs (no or primary treatment only) in the area
  tm_shape(wwEpPrimaryBathingLoc[grepl("^[01] ", wwEpPrimaryBathingLoc$TreatmentType),])+ 
  #    tm_text("RegNo", xmod = -2, ymod = -0.5, bgcol = "yellow", size=0.2) +
  
  # Finally add the problem WWTP discharge points
  tm_shape(wwEpPrimaryBathingLoc) +
  tm_symbols(shape = "wwtpAmrImpactLevel", size=0.4,
             legend.size.show = FALSE, 
             legend.col.show = FALSE,
             shapes = wwepsLevelsIcons) +
  
  # Impact zone
  tm_shape(st_buffer(cSegment,impactZoneBufferMetres)) +
  tm_borders(fill="#00000000", col="#00FFAA", lwd = 0.7, lty = "dashed") +
  
  # Health care facilities within sub-standard UWWTP agglomerations
  
  # Care homes.. there are a large number of these so only display
  # the top 5 in terms of bed capacity
  tm_shape(cHomesSubset[rev(order(cHomesSubset$Capacity))[1:5],]) + 
  tm_symbols(shape=tmap_icons("DATA/careHomeIcon.png"), size=0.18) +
  tm_text("Label", xmod = 0, ymod = 0.22, size=.45, alpha = 0.4, col = "#FF002990") +
  
  # Hospitals, again just add the top 5 in terms of bed capacity
  tm_shape(hospitalSubset[rev(order(hospitalSubset$hospitalBedCapacity))[1:5],]) + 
  tm_symbols(shape=tmap_icons("DATA/hospitalIcon.png"), size=0.22) +
  tm_text("Label", xmod = 0, ymod = 0.22, size=.45, alpha = 0.4, col = "#89512990") +
  
  # Bathingwater monitoring sites
  tm_shape(monSites) +
  tm_symbols(shape = "ecoliEnterococciCluster",
             title.shape = "Bathing water quality level:\nE.coli, Enterococci cluster centres",
             shapes = bathingLevelsIcons,
             size = 0.6) +
  tm_text("Name", xmod = 0, ymod = 0.22, size= .7, alpha = 0.3, col = "#89EE8890") +
  
  # WWTP
  # Ringsend WWTP
  tm_shape(ringsendWwtpSf) + tm_borders(col="red", lwd = 1) +
  
  # Also add in the primary emission point for Ringsend..
  tm_shape(wwEp[grepl("Ringsend", wwEp$Name) & wwEp$EmissionType=="Primary",]) + tm_dots(col="red", shape=4, lwd = 2,  size = 0.35) +
  
  # Scale bar
  tm_scale_bar(
    breaks = c(0, 2, 4, 6, 8, 10, 12),
    text.size = 0.4
  )  
```

Plot all available e.coli monitoring data for lowest water quality bathing locations in Dublin Bay.
```{r, fig.height = 8, fig.width = 7.5, fig.align = "left"}
ecoliWQoverview(epaBathingLocations, bqDataComplete, "North Bull Wall")

```


## Inland risk assessment

### AMR risk assessment case-study: Lake Muckno
```{r, fig.height = 8, fig.width = 7.5, fig.align = "left"}
tInlandStations = epaBqXyrsMediansStaged[match(inlandBathingWaters, epaBqXyrsMediansStaged$Code, nomatch=0),]

tInlandStations = st_as_sf(tInlandStations,
                           coords = c("Easting", "Northing"),
                           crs = 29903)

# Get the relevant monitoring sites
lkMno = tInlandStations[grep("Lough Muckno", tInlandStations$Name),]

# Get a list of all relevant waterbodies (lakes & rivers) withing a 10km radius for
# context (ie., Lough Muckno and tributaries etc.)
sfWfdLakeWaterBodies = st_as_sf(wfdLakeWaterBodies,
                                crs = 4326)
sfWfdLakeWaterBodies = st_transform(sfWfdLakeWaterBodies, crs=29903)


lakesSubset=unique(unlist(st_intersects(st_buffer(lkMno,10000),sfWfdLakeWaterBodies)))

# For rivers useuse Segment_Code to get a better perspective
riversSubset=sfWfdRiverWaterBodies[grepl("^06_", sfWfdRiverWaterBodies$Segment_Code,ignore.case = T),]

# Get a list of names for the map
riverNames = (sfWfdRiverWaterBodies[riversSubset,])[!is.na(sfWfdRiverWaterBodies[riversSubset,]$EPA_Name),]


# Transform industrial emission licensed (IEL) facilities feature collections to ITM
# (so we can use st_buffer() etc)
ielFacilities = st_transform(ielFacilities, crs=29903)

# Get the WWTP emission points to the lake
# For reference, get all WWTP emission points (including those with full tirtinary treatment)
epSubset=unique(unlist(st_intersects(st_buffer(sfWfdLakeWaterBodies[lakesSubset,],40),wwEp)))

# WWTP Agglomeration
wwtpAgglomSubset = wwtpAgglomerations[match(sub("-.*","",unique(wwEp[epSubset,]$RegNo)),wwtpAgglomerations$RegCd),]

# IEL

# Add human readable labels to industrial emission licensed (IEL) facilities.
ielSubset=unique(unlist(st_intersects(st_buffer(sfWfdLakeWaterBodies[lakesSubset,],10000),ielFacilities)))
ielSubset=ielFacilities[ielSubset,]
ielSubset$Type=sub("6.1 \\([ab]\\)", "Poultry Farm", ielSubset$LemaMajorClassOfActivity)
ielSubset$Type=sub("^11.5$", "Landfill", ielSubset$Type)
ielSubset$Type=sub("^5.12 \\(h\\)$", "Organic chemicals producer", ielSubset$Type)
ielSubset$Type=sub("^5.16$", "Biotec/Pharmaceuticals", ielSubset$Type)
ielSubset$Type=sub("^7.8 \\(a\\) \\(ii\\)$", "Food/feed treatment and processing", ielSubset$Type)
ielSubset$Type=sub("^7.2.1$", "Milk processing plant", ielSubset$Type)

# Set up a 5km impact zone and an 8.3km bounding box for some context.
impactZone = st_buffer(st_centroid(st_union(lkMno)), 5000)



# There are very few septic tanks in impact zone here so don't map them.
# Cattle density data is a more informative background in this case.
#  tm_shape(edSfSAP2022T6T7ED_2022)+
#  tm_fill("individualSepticTankPerKm",style = "cont", palette = "matplotlib.yl_or_rd", title = "Private household 2022 census,\nseptic tank numbers per km²") + 

#################################################
bbox_new <- st_bbox(st_buffer(st_centroid(st_union(lkMno)), 8300))

xrange <- bbox_new$xmax - bbox_new$xmin # range of x values
yrange <- bbox_new$ymax - bbox_new$ymin # range of y values

bbox_new[1] <- bbox_new[1] - (.1 * xrange) # xmin - left
bbox_new[3] <- bbox_new[3] + (.45 * xrange) # xmax - right
bbox_new[2] <- bbox_new[2] - (.1 * yrange) # ymin - bottom
bbox_new[4] <- bbox_new[4] + (.1 * yrange) # ymax - top

# Set the layout first
tm_layout(legend.outside = TRUE,legend.position = c("right", "top"),
          legend.outside.size = 0.5, 
          legend.text.size = 0.6,
          legend.title.size=0.6) +
  
  
  # Cattle densities
  tm_shape(edSfAVA42, bbox = bbox_new)+
  tm_fill("totalCattlePerKm",style = "cont", palette = "matplotlib.yl_or_rd", title = "CSO 2020 Animal census, cattle density per km²") + 
  
  # lake waterbodies
  tm_shape(sfWfdLakeWaterBodies[lakesSubset,]) + tm_fill(col = "#5599FFFF", fill="#5599FFFF") + 
  
  # Add rivers subset of interest at this site.
  tm_shape(sfWfdRiverWaterBodies[riversSubset,]) + tm_lines(col = "#5599FFFF", fill="#5599FFFF",lwd=2) +
  tm_shape(riverNames) + 
  #        tm_text("EPA_Name", xmod = 0, ymod = 0.22, size= .7, alpha = 0.3, col = "#89EE8890") +
  
  tm_shape(wwtpAgglomSubset) + tm_fill(col = "#559988FF", fill="#99FF8800", lwd=2, lty = "dashed") +
  tm_shape(castleblayneyUrbanEd) + tm_fill(col = "#559988FF", fill="#999988FF") +
  
  tm_shape(wwEp[epSubset,]) + tm_dots(col="red", shape=4, lwd = 1,  size = 0.15) +
  tm_text("EmissionType", xmod = 0, ymod = 0.22, size=.55, alpha = 0.2, col = "#54002990") +
  
  # Care homes (there are no major hospitals within agglomeration)
  #tm_shape(hospitalsSf) + 
  #tm_symbols(shape=tmap_icons("DATA/hospitalIcon.png"), size=0.22) +
  tm_shape(hiqaNursingHomes) + 
  tm_symbols(shape=tmap_icons("DATA/careHomeIcon.png"), size=0.3) + 
  tm_text("Label", xmod = 0, ymod = 0.22, size=.3, alpha = 0.2, col = "#FF002990") +
  
  # Add the WWTP outline
  # We use the one from OSM rather then that provided by EPA wwtpFacilities
  # as the OSM provides a better outline (wwtpFacilities is just a circle)
  tm_shape(castleblayneyWwtpSf) + tm_borders(col="red", lwd = 2) +
  
  # Bathingwater monitoring sites
  tm_shape(lkMno) +
  tm_symbols(shape = "ecoliEnterococciCluster",
             title.shape = "Bathing water quality level:\nE.coli, Enterococci cluster centres",
             shapes = tmap_icons(bathingLevelsIconsFiles[c(1,4,5)]),
             size = 1) +
  tm_text("Name", xmod = 0, ymod = 0.22, size= .6, alpha = 0.2, col = "#89EE8890") +
  
  # Add IEL facilities.
  tm_shape(ielSubset) + tm_symbols(shape=tmap_icons("ICONS/indEms.png"), size=1) +
  tm_text("Type", xmod = 0, ymod = 0.4, size= 0.5, alpha = 0.3, col = "#00000090") +
  
  tm_shape(impactZone) + tm_borders(fill="#00000000", col="#00FFAA", lwd = 2, lty = "dashed") +
  
  tm_scalebar() 


ecoliWQoverview(epaBathingLocations, bqDataComplete, "Lough Muckno Bathing Area")

# Driver densities around lake
driverDensitiesAroundWaterbody(st_centroid(st_union(lkMno)), 5000, "individualSepticTank", edSfSAP2022T6T7ED_2022)
driverDensitiesAroundWaterbody(st_centroid(st_union(lkMno)), 5000, "totalCattle", edSfAVA42)


```

### AMR risk assessment case-study: River Barrow stretch
```{r, fig.height = 8, fig.width = 7.5, fig.align = "left"}
tInlandStations = epaBqXyrsMediansStaged[match(inlandBathingWaters, epaBqXyrsMediansStaged$Code, nomatch=0),]

tInlandStations = st_as_sf(tInlandStations,
                           coords = c("Easting", "Northing"),
                           crs = 29903)


# Get the relevant monitoring sites
barrowSites = tStations[grep("Barrow", tStations$Name),]

# Estimate the impact zone.
# Draw a circular buffer around the upstream monitoring site with a radius
# slightly greater than the distance between the two sites.
impactZone = st_buffer(barrowSites[2,], 1.2*st_distance(barrowSites[1,], barrowSites[2])[2])

# WWTP EPs
# There are no discharge points from large (>500 PE) substandard wastewater treatment plants (WWTPs)
# within the defined impact zone.
# Given the numerous tributaries of the River Barrow within the study area, all relevant
# WWTP discharge points, including those providing tertiary-level treatment, are examined
# here for completeness.

# FYI: Carlow WWTP also acts as a disposal site for leachate from the Powerstown Landfill.
epSubset=unique(unlist(st_intersects(impactZone,wwEp)))

# Small WWTPs (pop<500) operating under the EPA small discharge license (license prefix 'A')
# have limited info in EPA data download portal.
# Also 2 large WWTPs have an outdated ActiveLicenceNumber listed in the WWTP facilities data
# from the EPA resulting in missing treatment type and PE data. This is corrected here also.
#
# The following data was added manually by comparing with
# licence and compliance information from the LEAP portal.
wwEpBarrowSubset=wwEp[epSubset,]
entriesRequiringUpdate = is.na(wwEpBarrowSubset$TreatmentType)
wwEpBarrowSubset[entriesRequiringUpdate,]$TreatmentType = c("2 - Secondary Treatment", "2 - Secondary Treatment", "2 - Secondary Treatment", "2 - Secondary Treatment", "2 - Secondary Treatment", "2 - Secondary Treatment", "2 - Secondary Treatment", "1 - Primary Treatment", "2 - Secondary Treatment", "1 - Primary Treatment", "1 - Primary Treatment", "1 - Primary Treatment", "2 - Secondary Treatment")

wwEpBarrowSubset[entriesRequiringUpdate,]$wwtpPlantDesignPE = c(100, 120, 200, 250, 200, 499, 50, 266, 27, 75, 65, 21, 60)

# Only display EPs discharging directly into river Barrow tributaries.
epSubset=unique(unlist(st_intersects(sfWfdRiverWaterBodies[grepl("IE_SE_14", sfWfdRiverWaterBodies$River_Waterbody_Code,ignore.case = T),],st_buffer(wwEpBarrowSubset,100))))
epSubset=wwEpBarrowSubset[epSubset,]

# There is one small WWTP providing primary treatment only, serving a
# population equivalent (PE) of 50. Given its small scale and its distance
# from the bathing water monitoring sites, it is not expected to significantly
# influence bathing water quality results.
# epSubset=epSubset[epSubset$TreatmentType=="1 - Primary Treatment",]

# Now that we've ruled out all smaller WWTPs, to avoid clutter
# we just map all major WWTPs on this stretch of the Barrow.
epSubset=epSubset[epSubset$wwtpPlantDesignPE>1000,]

# Add a label for the map
epSubset$Label = paste0(epSubset$Name,", ",epSubset$wwtpAgglomerationPE)

# IELs: Add human readable labels for indistrial emissions license numbers.
# see https://epawebapp.epa.ie/licences/lic_eDMS/090151b280885849.pdf
# https://www.epa.ie/publications/licensing--permitting/industrial/ied/online-application-form---attachment-741-emissions-to-air-main.php#:~:text=Online%20Application%20Form%20%2D%20Attachment%207.4.%201,to%20Air%20Main%20%7C%20Environmental%20Protection%20Agency.
#https://leap.epa.ie/docs/a7d3b2a9-64c2-4608-8a72-2a92619ae53b.pdf

# Also, for landfill note,
# https://epawebapp.epa.ie/licences/lic_eDMS/090151b280514249.pdf
# https://www.sciencedirect.com/science/article/pii/S2214785323022599

ielSubset=unique(unlist(st_intersects(impactZone,ielFacilities)))
ielSubset=ielFacilities[ielSubset,]
ielSubset$Type=sub("6.1 \\([ab]\\)", "Poultry Farm", ielSubset$LemaMajorClassOfActivity)
ielSubset$Type=sub("6.2 \\(a\\)", "Pig Farm", ielSubset$LemaMajorClassOfActivity)
ielSubset$Type=sub("^11.5$", "Landfill", ielSubset$Type)
ielSubset$Type=sub("^11.4 \\(b\\)\\(i\\)$", "Waste Composting or Anaerobic Digestion >75 t/day", ielSubset$Type)
ielSubset$Type=sub("^5.12 \\(h\\)$", "Organic chemicals producer", ielSubset$Type)
ielSubset$Type=sub("^5.16$", "Biotec/Pharmaceuticals", ielSubset$Type)
ielSubset$Type=sub("^7.8 \\(a\\) \\(ii\\)$", "Animal/vegetable raw materials processing > 300 tonnes per day", ielSubset$Type)
ielSubset$Type=sub("^7.8 \\(a\\) \\(iii\\)$", "Animal/vegetable raw materials processing > 75 tonnes per day", ielSubset$Type)
ielSubset$Type=sub("^7.2.1$", "Milk processing plant", ielSubset$Type)
ielSubset$Type=sub("^7.4*", "Plant with emission to air", ielSubset$Type)

# Add a more detailed description of what each plant does.
ielSubset$Description = ielSubset$RegCD
ielSubset$Description=sub("^W0025$", "Closed landfill, ~2,200 m3 leachate removed annually to Mortarstown Waste Water Treatment Plant", ielSubset$Description)
ielSubset$Description=sub("^P0961$", "Pig farm, >6000 adult, >4000 weaners", ielSubset$Description)
ielSubset$Description=sub("^P0846$", "Meat processing plant", ielSubset$Description)
ielSubset$Description=sub("^P0186$", "Closed abbatoir/meat processing 2004", ielSubset$Description)
ielSubset$Description=sub("^W0284$", "Bulk composting/food recycling/biowaste &
sewage sludge >75 t/day", ielSubset$Description)
ielSubset$Description=sub("^P0825$", "Pig farm, >7000 adult, >6000 weaners", ielSubset$Description)
ielSubset$Description=sub("^P1069$", "Animal feed (equine/agri/pet)", ielSubset$Description)
ielSubset$Description=sub("^P0959$", "Pig farm, licence ceased (never commenced)", ielSubset$Description)
ielSubset$Description=sub("^P0959$", "Pig farm, licence ceased (never commenced)", ielSubset$Description)
ielSubset$Description=sub("^P0400$", "Limestone mining and processing", ielSubset$Description)
ielSubset$Description=sub("^P0556$", "Timber sawmills", ielSubset$Description)
ielSubset$Description=sub("^P0759$", "Lime production, license ceased", ielSubset$Description)
ielSubset$Description=sub("^P0931$", "R&D, gas-fired peaking power plants", ielSubset$Description)

# Remove IELs that are not relevant to our analysis (no longer in operation, not relevant to environmental AMR etc.).
ielSubset = ielSubset[-match(c("P0931-01","P0759-01","P0556-01","P0959-01","P0186-01"),ielSubset$ActiveLicenceNumber),]
ielSubset = ielSubset[!duplicated(ielSubset$ActiveLicenceNumber),]

#ielSubset[,c("Type","Name", "ActiveLicenceNumber")]

# Get any surrounding rivers that intersect or run close by.
riversSubset=unique(unlist(st_intersects(impactZone,sfWfdRiverWaterBodies)))


# Draw map
bbox_new <- st_bbox(st_buffer(impactZone, 2000))
xrange <- bbox_new$xmax - bbox_new$xmin # range of x values
yrange <- bbox_new$ymax - bbox_new$ymin # range of y values

bbox_new[1] <- bbox_new[1] - (.1 * xrange) # xmin - left
bbox_new[3] <- bbox_new[3] + (.13 * xrange) # xmax - right
bbox_new[2] <- bbox_new[2] - (.1 * yrange) # ymin - bottom
bbox_new[4] <- bbox_new[4] + (.1 * yrange) # ymax - top

# Set the layout first
tm_layout(legend.outside = TRUE,legend.position = c("right", "top"),
          legend.outside.size = 0.5, 
          legend.text.size = 0.6,
          legend.title.size=0.6) +
  
  # Cattle
  tm_shape(edSfAVA42, bbox = bbox_new) +
  tm_fill("totalCattlePerKm",style = "cont", palette = "matplotlib.yl_or_rd", title = "CSO 2020 Animal census, cattle density per km²") + 
  
  # Septic tanks  
  # tm_shape(edSfSAP2022T6T7ED_2022, bbox = bbox_new) +
  #  tm_fill("individualSepticTankPerKm",style = "cont", palette = "matplotlib.yl_or_rd", title = "Private household 2022 census,\nseptic tank numbers per km²") +
  
  # Carlow town
  tm_shape(carlow) + tm_fill(col = "#559988FF", fill="#999988FF") +
  tm_shape(graigue) + tm_fill(col = "#559988FF", fill="#999988FF") +
  tm_shape(graigueUrbanED) + tm_fill(col = "#559988FF", fill="#999988FF") +
  
  # Leighlinbridge
  tm_shape(leighlinbridge) + tm_fill(col = "#559988FF", fill="#999988FF") +
  
  # Bagenalstown
  tm_shape(bagenalstown) + tm_fill(col = "#559988FF", fill="#999988FF") +
  
  # Impact zone  
  tm_shape(impactZone) + tm_lines(col = "red", fill="#5599FF33",lwd=2) +
  
  
  # Barrow catchment: hydrometric area 14
  tm_shape(sfWfdRiverWaterBodies[riversSubset,]) + tm_lines(col = "#5599FFFF", fill="#5599FFFF",lwd=1) +
  
  tm_shape(sfWfdRiverWaterBodies[grepl("IE_SE_14", sfWfdRiverWaterBodies$River_Waterbody_Code,ignore.case = T),]) + tm_lines(col = "#5577FFFF", fill="#5577FFFF",lwd=2) +
  
  tm_shape(sfWfdRiverWaterBodies[grepl("Barrow", sfWfdRiverWaterBodies$EPA_Name,ignore.case = T),]) + tm_lines(col = "#5577FFFF", fill="#5577FFFF",lwd=4) +
  
  # Barrow bathingwater monitoring sites
  tm_shape(tStations[grep("Barrow", tStations$Name),]) +
  tm_symbols(shape = "ecoliEnterococciCluster",
             title.shape = "Bathing water quality level:\nE.coli, Enterococci cluster centres",
             shapes = tmap_icons(bathingLevelsIconsFiles[c(4,5)]),
             size = 1) +
  tm_text("Name", xmod = 0, ymod = 0.22, size= .6, alpha = 0.3, col = "#89EE8890") +
  
  
  # Add IEL facilities.
  tm_shape(ielSubset) + tm_symbols(shape=tmap_icons("ICONS/indEms.png"), size=1) +
  tm_text("Description", xmod = 0, ymod = 0.4, size= 0.6, alpha = 0.3, col = "#00000090") +
  
  
  tm_shape(epSubset) + tm_dots(col="red", shape=4, lwd = 2,  size = 0.6) +    tm_shape(epSubset[epSubset$EmissionType=="Primary",]) +
  tm_text("Label", xmod = 0, ymod = 0.22, size=.4, alpha = 0.2, col = "#54002990") +
  
  # Add relevant EPs for D0090
  tm_shape(epSubset[grepl("TPEFF0100D0090", epSubset$EmissionID),]) +
  tm_text("EmissionID", xmod = 0, ymod = 0.22, size=.4, alpha = 0.2, col = "#6600FF90") +
  
  
  # Scale bar
  tm_scalebar(breaks = c(0, 2, 4, 6, 8,10,12,14,16),text.size = 0.4)  

# Cattle densities
# Densities inside impact zone.
driverDensitiesAroundWaterbody(barrowSites[1,], 2000, "totalCattle", edSfAVA42)


# As a percentage of the national average
driverDensitiesAroundWaterbody(barrowSites[2,], (1.2*st_distance(barrowSites[1,], barrowSites[2])[2]), "totalCattle", edSfAVA42)/driverDensitiesAroundWaterbody(barrowSites[2,], 500000, "totalCattle", edSfAVA42)

# Septic tank densities
# Densities inside impact zone as a percentage of the national average
driverDensitiesAroundWaterbody(barrowSites[2,], (1.2*st_distance(barrowSites[1,], barrowSites[2])[2]), "individualSepticTank", edSfSAP2022T6T7ED_2022)/driverDensitiesAroundWaterbody(barrowSites[2,], 500000, "individualSepticTank", edSfSAP2022T6T7ED_2022)

# Plots
# ecoliWQoverview(epaBathingLocations, bqDataComplete, "River Barrow Bagenalstown")
```

